---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Respiratory Rate Determination by Non-Invasive Means

This repository consists of an R project for COMP90072 - The Art of Scientific Computation, holding the source files for reproducing the deliverable, including the code files and their version history.

## Requirements

To reproduce the deliverable of the project, you will need to install the following packages and their dependencies as follows:

```{r lib, eval = FALSE}
install.packages("tidyverse")
install.packages("zoo")
install.packages("patchwork")
install.packages("tinytex")
tinytex::install_tinytex()
```

## Reproducible research

The deliverable of the project consists of a PDF report rendered by RMarkdown.

```{r render, eval = FALSE}
rmarkdown::render("report/report.Rmd")
```

The [report](https://github.com/szmsu2011/comp90072/tree/main/report) directory consists of source files for reproducing the PDF report. The [R](https://github.com/szmsu2011/comp90072/tree/main/R) directory consists of source files for the R functions and workflow caches. The [data-bin](https://github.com/szmsu2011/comp90072/tree/main/data-bin) directory holds the binary data files for the project.
