---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(width = 100)
```

# Respiratory Rate Determination by Non-Invasive Means

This repository consists of an R project for COMP90072 - The Art of Scientific Computation, holding the source files for reproducing the deliverables, including the code files and their version history.

## Requirements

To reproduce the deliverables of the project, you will need to install the following packages and their dependencies as follows:

```{r lib, eval = FALSE}
install.packages("xaringan")
install.packages("xaringanthemer")
install.packages("tidyverse")
install.packages("furrr")
install.packages("zoo")
```

## Reproducible research

The deliverables of the project consist of a PDF report and an HTML slide rendered by RMarkdown.

```{r render, eval = FALSE}
rmarkdown::render("report/report.Rmd")
# rmarkdown::render("slide/slide.Rmd")
```
